The CMS Data Aggregation System

### 2010-05-04 doc/paper.tex
Work on Section 3, 4, 4.1
git-svn-id: svn+ssh://svn.cern.ch/reps/CMSDMWM/DAS/trunk@1806 4525493e-7705-40b1-a816-d608a930855b

\section{CMS data model\label{DataModel}}
The CMS, Compact Muon Solenoid, \cite{CMS} 
is one of the two large general-purpose particle physics detectors built on 
the proton-proton Large Hadron Collider (LHC) at CERN in Switzerland and France. 
It is designed to explore the frontier of the HEP and provide physicists
the ability to look at conditions presented at the early stage of our Universe.
More then 3000 physicists from 183 institutions representing almost 
38 countries are involved in the design, construction and maintenance of the experiment.

The CMS distributed computing and data model \cite{CMSDataModel} 
is designed to process
and efficiently manage a few PBs of data expected each year
in CMS operation at LHC. The computing resources provided by members of CMS
collaboration are geographically distributed, 
interconnected via high throughput networks and operated by means 
of Grid software. The model allows to cover broad variety of
hardware, mass and storage elements and configuration of the
clusters. To accommodate efficient data processing CMS uses
a multi-Tier distributed model, where specific tasks of data taking,
processing, archival and distribution are assigned to each tier based
on CMS data model. For example, the Tier-0 center at CERN is responsible
for archiving the data coming out from the detector, prompt first pass reconstruction
and data distribution to Tier-1 centers, located around the world.
%The 7 Tier-1 centers
%located in France, Germany, Italy, Spain, Taiwan, United Kingdom and United States
%keep a portion (copy) of the data delivered by Tier-0 for further processing.
They provide storage and CPU power for high priority analysis jobs.
The Tier-2 centers
%, located at more then 50 sites around the world,
are dedicated for user analysis tasks and production of simulated data.

@@

Once such conglomerate of data-services starts operating an obvious
question arise: how to find out desired information across multiple data-services
in our distributed environment? Even though individual data-services were designed
to answer specific questions about the data they serve, the ability to search and relate
information among them was always a tedious human task. The growing amount of information
and desire to make cross-service queries force us to design and develop new
Data Aggregation System.

\section{Data Aggregation System\label{DAS}}
The goal of the Data Aggregation System was to combine and
aggregate meta-data from existing data-services
under one umbrella, utilizing their access rules, APIs, security policies, etc., and
allowing users to perform queries against them. In contrast, the approaches
discussed in \cite{DBXplorer, QueryAnswer, FedDB} are
enforced a publication step of data into a dedicated database 
or central repositories \cite{iRODS}. Therefore, quite often it is required
that data should be converted into uniform data formats in order enable
aggregation, see \cite{OpenArchive}. In DAS architecture the data are provided
by existing data-services and infrastructure, whose implementation,
data representation and security policies are left intact. Consequently,
DAS does not require a data preservation and transaction capabilities. 
Instead DAS can be treated
as proxy or caching layer where information retrieval are triggered dynamically
by user queries. Such infrastructure allows data sharing across different queries and
data aggregation using appropriate mapping. Below we discuss in details the
DAS architecture.

\subsection{DAS architecture}
The design of the DAS system was based on previous studies of the CMS Data 
Bookkeeping System (DBS) \cite{DBS, DBS07}. The DBS Data Discovery service
\cite{DD} was based on DBS Query Language (DBS-QL) \cite{DBS-QL}. It provided
a syntax similar to SQL, without specific knowledge of underlying DB schema.
%We carefully analyzed user
%requests and queries placed into the DBS, their patterns, frequencies and latency. 
%As a result we created a data discovery service \cite{DD}. Originally it used MVC
%architecture and later, being re-factored as pure presentation layer on top of 
%the DBS Query Language (DBS-QL) \cite{DBS-QL} with support of dynamic queries against
%DBS back-ends. The DBS-QL \cite{DBS-QL} syntax was based on SQL query language, where 
%we took away relations and details on underlying schema from end-users. 
%This was achieved by using Dijkstra's shortest path algorithm 
%to establish necessary table joins based on provided information in user queries and
%knowledge of foreign key relationships from auto-loaded database schema.
%The DBS-QL keys exposed to end-users were chosen based on a jargon used by
%physicists. They're internally mapped into database entities and attributes, 
%e.g. the {\it run} DBS-QL key refers to run number attribute of run summary table.
A quick adoption and wide usage of DBS-QL in CMS gave us confidence in chosen approach 
and provide a basis for building DAS system.

As was pointed out in \cite{Arms} a mixed content and 
mixed meta-data and meta-data consistency should be considered as a whole in design 
of the system to successful information discovery. 
Starting from this ground we designed DAS system as an
additional layer on top of the existing data-services
within CMS computing infrastructure by imposing the following set of requirements:
\begin{itemize}
\item support a keyword based search queries with ability to use condition operators;
%\item pluggable interface to existing and up-coming data-service APIs;
\item use existing heterogeneous software environment and distributed nature of the data-services;
\item preserve access rules and security policies of individual data-services;
%\item be flexible to adjust to dynamic nature of data-service APIs, 
%due to continuously changing detector and software environments;
\item retrieve data on demand and aggregate them if necessary across
multiple data-services;
\item be transparent to data-services implementation and their data formats;
%\item support legacy data-services, applications and APIs.
\end{itemize}
\begin{figure}[htb]
\centering
\includegraphics[width=100mm]{DAS_Cache_and_Analytics.pdf}
\caption{
The DAS architecture diagram. 
Input query were placed into analytics DB and mapped into a set of
data-service APIs and their parameters. If superset of the query has been
found in DAS cache, the appropriate results were delivered back to the users.
Otherwise a data-service APIs were invoked, triggering DAS cache population. 
Once data become available and aggregated in DAS cache, users were able 
to see the results. This on demand system was supplemented by DAS robots 
(Unix daemons) who were consulting DAS analytics database to enable data 
population into DAS cache for most common query requests.
}
\label{DAS_cache}
\end{figure}

\noindent
Based on this set we rule out the choice of RDMS as DAS back-end(s)
for several reasons. We agreed to support dynamic queries, data
aggregation on demand and pluggable data-services. Therefore a choice 
of concrete schema for data storage was problematic. Moreover we didn't impose
any transactions and data persistence in DAS. This allows us to look 
for alternative IT solutions.
Through the analysis of available options, such as a file-based and memory caches, 
key-value databases, documented-oriented databases we made our choice in favor 
of the last technology. Among them we evaluated CouchDB \cite{CouchDB} and 
MongoDB \cite{MongoDB}. Both systems provide schema free document
storage, replication and fail-over features, but our choice in favor of 
MongoDB was obvious, due to its support of dynamic queries, 
full indexes, including inner objects and embedded arrays,
and auto-sharding. Our preliminary benchmarks shown that it can sustain
the desired load and size of our meta-data information. We used 
document-oriented database, MongoDB, as a back-end for the three 
main DAS components: Mapping, Analytics and DAS cache databases, 
which will be discussed next.

The DAS architecture is shown in Fig. \ref{DAS_cache}. It consists of
the core library with support of pluggable modules for data retrieval;
caching layer to store data-service output and aggregate results;
request service to handle user queries;
Mapping DB to keep information about data-service APIs, their
notations and mapping to DAS;
Analytics DB for query analysis and identification of pre-fetch 
strategies.

>>>>

\section{CMS data model\label{DataModel}}
The CMS, Compact Muon Solenoid, \cite{CMS} 
is one of the two particle physics detectors built on 
the proton-proton Large Hadron Collider (LHC) at CERN in Switzerland and France. 
It is designed to explore the frontier of the HEP and provide physicists
the ability to look at the conditions presented at the early stage of our Universe.
More then 3000 physicists from 183 institutions representing 38 countries 
are involved in the design, construction and maintenance of the experiment.

The CMS distributed computing and data model \cite{CMSDataModel} 
is designed to process
and efficiently manage a few PBs of data expected each year
in CMS operation at LHC. The computing resources provided by members of CMS
collaboration are geographically distributed, 
interconnected via high throughput networks and operated by means 
of the Grid software. The model allows to cover broad variety of
hardware, mass and storage elements and configuration of the
clusters. To accommodate efficient data processing CMS uses
a multi-Tier distributed model, where specific tasks of data taking,
e.g. processing, archival and distribution, are assigned to each tier based
on CMS data model. For example, the Tier-0 center at CERN is responsible
for archiving the data coming out from the detector, prompt first pass reconstruction
and data distribution to Tier-1 centers, located around the world.
%The 7 Tier-1 centers
%located in France, Germany, Italy, Spain, Taiwan, United Kingdom and United States
%keep a portion (copy) of the data delivered by Tier-0 for further processing.
They provide storage and CPU power for high priority analysis jobs.
The Tier-2 centers
%, located at more then 50 sites around the world,
are dedicated for user analysis tasks and production of simulated data.

@@

Once such conglomerate of data-services starts operating the obvious
question arises: how to find out desired information across multiple data-services
in our distributed environment? Even though individual data-services were designed
to answer specific questions about the data they serve, the ability to search and relate
information among them was always a tedious human task. The growing amount of information
and desire to make cross-service queries force us to design and develop new
Data Aggregation System.

\section{Data Aggregation System\label{DAS}}
The goal of the Data Aggregation System was to fetch and aggregate meta-data 
on demand from existing CMS data-services under one umbrella by using their APIs, 
security policies, etc.
In contrast, the approaches
discussed in \cite{DBXplorer, QueryAnswer, FedDB} are
enforced a publication step of data into a dedicated database 
or central repositories \cite{iRODS}. Therefore, quite often it was required
that data should be converted into common data structure/format in order to enable
aggregation, see \cite{OpenArchive}. In DAS the data are provided
by existing data-services and infrastructure, whose implementation,
data representation and security policies are left intact. Consequently,
DAS does not require a data preservation and transaction capabilities. 
Instead DAS can be treated as proxy or caching layer where information 
retrieval and aggregation are triggered dynamically by user queries. 
Below we discuss in details the DAS architecture.

\subsection{DAS architecture}
The design of the DAS system was based on previous studies of CMS Data 
Bookkeeping System (DBS) \cite{DBS, DBS07}. The DBS Data Discovery service
\cite{DD} was based on DBS Query Language (DBS-QL) \cite{DBS-QL}. It provided
a syntax similar to SQL, without specific knowledge of underlying DB schema.
%We carefully analyzed user
%requests and queries placed into the DBS, their patterns, frequencies and latency. 
%As a result we created a data discovery service \cite{DD}. Originally it used MVC
%architecture and later, being re-factored as pure presentation layer on top of 
%the DBS Query Language (DBS-QL) \cite{DBS-QL} with support of dynamic queries against
%DBS back-ends. The DBS-QL \cite{DBS-QL} syntax was based on SQL query language, where 
%we took away relations and details on underlying schema from end-users. 
%This was achieved by using Dijkstra's shortest path algorithm 
%to establish necessary table joins based on provided information in user queries and
%knowledge of foreign key relationships from auto-loaded database schema.
%The DBS-QL keys exposed to end-users were chosen based on a jargon used by
%physicists. They're internally mapped into database entities and attributes, 
%e.g. the {\it run} DBS-QL key refers to run number attribute of run summary table.
A quick adoption and wide usage of DBS-QL in CMS gave us confidence in chosen approach 
and provide a basis for building DAS system.

As was pointed out in \cite{Arms} a mixed content and 
mixed meta-data and meta-data consistency should be considered as a whole in design 
of the system to successful information discovery. 
Starting from this ground we are designed DAS system as an
additional layer on top of the existing data-services
within CMS computing infrastructure by imposing the following set of requirements:
\begin{itemize}
\item support a keyword based search queries with ability to use condition operators;
%\item pluggable interface to existing and up-coming data-service APIs;
\item use existing heterogeneous software environment and distributed nature of the data-services;
\item preserve access rules and security policies of individual data-services;
%\item be flexible to adjust to dynamic nature of data-service APIs, 
%due to continuously changing detector and software environments;
\item retrieve data on demand and aggregate them if necessary across
multiple data-services;
\item be transparent to data-services implementation and their data formats;
%\item support legacy data-services, applications and APIs.
\end{itemize}
\begin{figure}[htb]
\centering
\includegraphics[width=100mm]{DAS_Cache_and_Analytics.pdf}
\caption{
The DAS architecture diagram. 
Input query were placed into analytics DB and mapped into a set of
data-service APIs and their parameters. If superset of the query has been
found in DAS cache, the appropriate results were delivered back to the users.
Otherwise a data-service APIs were invoked, triggering DAS cache population. 
Once data become available and aggregated in DAS cache, users were able 
to see the results. This on demand system was supplemented by DAS robots 
(Unix daemons) who were consulting DAS analytics database to enable data 
population into DAS cache for most common query requests.
}
\label{DAS_cache}
\end{figure}

\noindent
The choice of existing RDMS as DAS back-end(s) was ruled out for several reasons. 
For instance, we didn't require any transactions and data persistency in DAS;
a dynamically typing of stored meta-data objects was one of the requirements, etc.
Those arguments forces us to look for alternative IT solutions.
Through the careful analysis of available options, such as a file-based and in memory caches, 
key-value databases, documented-oriented databases we made our choice in favor 
of the last technology. Among them we evaluated CouchDB \cite{CouchDB} and 
MongoDB \cite{MongoDB}. Both systems are ``schemaless'' databases providing
arbitrary document structure storage, replication and fail-over features. 
We made our decision in favor of MongoDB, due to its support of dynamic queries, 
full indexes, including inner objects and embedded arrays,
and auto-sharding. Our preliminary benchmarks shown that it can sustain
the desired load and size of our meta-data information. We used 
document-oriented database, MongoDB, as a back-end for three 
main DAS components: Mapping, Analytics and DAS cache databases, 
which will be discussed next.

The DAS architecture is shown in Fig. \ref{DAS_cache}. It consists of
the core library with support of pluggable modules for data retrieval;
caching layer to store and aggregate results;
request service to handle user queries;
Mapping DB to keep information about data-service APIs, e.g.
notations and mapping to DAS keys;
Analytics DB for query analysis.
